\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-21 honors_thesis/ppsn_paper/paper.tex
Finished PPSN version of the paper and submitted it

%\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
\author{Author names omitted for review}

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{chu-2006-intelligent-parallel-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

We show that the results of standard PSO can be reproduced \emph{exactly}, two
iterations at a time, using a speculative approach similar to speculative
execution. We prove that the standard PSO equation can be factored such that a
set of speculative positions can be found which will \emph{always} include the
position computed in the next iteration.  By computing the value of the
objective function for each of the speculative positions at the same time the
algorithm evaluates the objective function for the current position, it is
possible to know the objective function values for both the current and the
next iteration at the same time.  The resulting implementation runs efficiently
on large clusters where the number of processors is much larger than a typical
or reasonable number of particles, producing better results in less
``wall-clock'' time.

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once.
\sec{extra} then introduces a variant of the speculative algorithm that in some
cases performs better.  In \sec{results} and \sec{conclusion} we present our
results and conclude.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-2002-constricted-pso}.

@@

Although this speculative approach computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
ignored, and might just as well never have been computed.  We call the set of
computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

@@

In order to do two iterations at once, we use 8 times as many processors as
there are particles in the swarm.  If these processors were not performing
speculative evaluation, they might instead be used for function evaluation
needed to support a large swarm.  This raises the question of whether a swarm
of 100 particles doing twice as many iterations outperforms a swarm of 800
particles.  We show in the rest of this paper that in many, instances, a
smaller swarm performing more iterations does in fact outperform a larger
swarm.  We acknowledge that both intuition and prior
research~\cite{mcnabb-2009-large-particle-swarms} indicate that the
optimization of deceptive functions benefits greatly from large and even very
large swarm sizes. Thus this work will focus on less deceptive functions.

@@

A particle at iteration $t-1$ that has been moved to iteration $t$ using
\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has not yet
been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has been
evaluated, but it has still not yet received information from its neighbors, it
is denoted as $\nonbest{\p}_t$.  Only when the particle has updated its
neighborhood best is it a complete particle at iteration $t$.  It is then
simply denoted as $\p_t$.

@@

\section{Using All Speculative Evaluations}
\label{sec:extra}

In performing speculative evaluation as we have described it, $2n+1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of the
information obtained through those evaluations instead of ignoring it
Making use of this information changes the
behavior of PSO, instead of reproducing it exactly as the above method
explains, but the change turns out to be an improvement in our context.

To make better use of the speculative evaluations, instead of choosing the speculative
child that matches that branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The branch that PSO would have taken is ignored, but $\p_t$ must still be
created so it can give the right personal best and neighborhood best values and
positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch. We call this
varient ``Pick Best''.

This can be thought of as drawing a number of samples from the next iteration
and accepting the best one.  Speculative particles that move in good directions
are kept.  It seems this modification to PSO would contribute local exploration,
having knowledge of all of the speculative children, but more global exploitation,
since Pick Best only considers the best of these alternatives.
With functions that are less deceptive, or only locally deceptive,
Pick Best works well, while performance on more deceptive
functions would probably be hurt.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
evaluations.  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of at least seconds (or more) to
evaluate; in such cases our assumption is perfectly reasonable.

In each set of experiments we keep the number of processors for each algorithm
constant.  In all of our SEPSO experiments each particle has two neighbors in
addition to itself.  As shown in Table~\ref{tab:evals}, this results in $7$
speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have a total of $8p_s$ evaluations for
every two iterations, where $p_s$ is the number of particles in the speculative
swarm.  We run these experiments on $8p_s$ processors.  In the base-line
experiments, standard pso the same number of processors and thus enjoys the
benefit of $8p_s$ particles running on $8p_s$ processors.  At least in
benchmark functions, extra particles only accelerate the performance of
PSO~\cite{mcnabb-2009-large-particle-swarms}. They generally have diminishing
value, as seen earlier in \fig{evals-sphere}, but these extra particles give
standard PSO the best possible context and full access to the same resources as
SEPSO.

As discussed in \sec{implementation},
where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion of speculative
evaluations.
For the standard PSO baseline we have included experiments with the
ring and the random topologies, both with two neighbors, as well as for the complete topology.
It is important to note however, that in many cases the complete topology is not a
practical alternative in the context of large
swarms on large cluster where where the messaging complexity is $O(n^2)$ and can
overwhelm the system.

We follow the experimental set up proposed by Bratton, et.al.~\cite{bratton-2007-defining-a-standard-for-pso}. 
All functions have 20 dimensions, and all results shown
are averages over 20 runs.  We use 240 processors in our experiments, so
speculative algorithms use 30 particles and standard PSO uses 240 particles.
For ease of labeling, we call our speculative algorithm SEPSO, the variant
SEPSO Pick Best, and standard PSO just PSO and identify the topology in a suffix,
``PSO Ring'' for example.

The benchmark function Sphere ($f(\Vec{x}) = \sum_{i=1}^D x_i^2$)
has no local optima and is most efficiently optimized using
a small swarm but many iterations.
In \fig{sphere1}, we can see
that SEPSO clearly beats PSO with a Random topology or a ring topology.

SEPSO approaches the performance of PSO with the complete topology, even though
PSO with the complete topology requires $O(n^2)$ communication.
SEPSO Pick Best handily outperforms all other algorithms
in this problem.

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{sphere1}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round. Note that PSO Complete requires $O(n^2)$ messaging
  and may not be practical in many cases.}
  \label{fig:sphere1}
\end{figure}

The benchmark function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally best solved in pso using the Ring topology,
as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it does not.  Instead of
showing average function value at each iteration, a more enlightening plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration~\cite{bratton-2007-defining-a-standard-for-pso}.

PSO and SEPSO get caught in local minima with small swarm sizes so
we show results in \fig{griewank1} for swarms of size 100 (SEPSO) and 800 (standard)
using the Ring topology 
\fig{griewank2} shows that SEPSO quickly finds
the global optimum essentially 100\% of the time.

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank2}
\end{figure}

As noted in the introduction previous work has shown that the optimization of highly
deceptive functions like Rastrigin
($f(\Vec{x}) = \sum_{i=1}^D \left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$)
benefit greatly from the addition of particles. 
Smaller swarms get caught
in local optima.  Because our speculative algorithms have a significantly
smaller swarm size, they get stuck at higher values while the larger swarms
performing regular PSO continue to improve the best value found.
Our experiments with SEPSO on Rastrigin were predicable lack luster,
yielding a vaerage value of xxx after x evaluations, as compared to y for 
standard PSO.

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using our
modifications, the original PSO algorithm is exactly reproduced two iterations
at a time.  This technique requires more function evaluations per iteration
than regular PSO, but for some functions still performs better when run in a
parallel environment.

We have also described a method for making use of extra speculative evaluations
that performs incredibly well on some functions.

Though space has limited our presentation of results to a very small sample of
the interesting and relevant applications of our technique, we believe the
results shown are sufficient motivation for the further exploration of
speculative evaluation in PSO.

Our results show that performing speculative evaluation instead of increasing
swarm size is only beneficial for some functions at the swarm sizes we used.
However, we would venture that for all functions there is some point at which
increasing the swarm size provides relatively little benefit.  At that point,
additional processors should be used to speed up the algorithm instead of add
additional particles.

There are some functions for which very little exploration needs to be done;
Sphere is an example of such a function.  For such functions the best use of
processors is to have a small swarm performing speculative evaluation with our
second method, where all speculative evaluations are used.

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, 30
particles is not ``enough'' exploration, but 100 is.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-2009-large-particle-swarms}.  For those functions the correct
use of processors cannot be easily prescribed and depends largely on the
desired outcome and how many processors are available.  It is likely that the
best thing to do with those functions is to use all processors to add particles
to the swarm and do no speculative evaluation.

>>>>

\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}
\institute{Department of Computer Science, Brigham Young University}

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{belal-2004-parallel-models-for-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{mostaghim-2006-multi-objective-pso-on-grids,
venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

We show that the results of standard PSO can be reproduced \emph{exactly}, two
iterations at a time, using a speculative approach similar to speculative
execution. We prove that the standard PSO equations can be factored such that a
set of speculative positions can be found which will \emph{always} include the
position computed in the next iteration.  By computing the value of the
objective function for each of the speculative positions at the same time the
algorithm evaluates the objective function for the current position, it is
possible to know the objective function values for both the current and the
next iteration at the same time.  The resulting implementation runs efficiently
on large clusters where the number of processors is much larger than a typical
or reasonable number of particles, producing better results in less
``wall-clock'' time.

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once.  In
\sec{results} and \sec{conclusion} we present our results and conclude.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen.  The parameters \( \ncoeff \), \( \pcoeff \), and \( \constriction
\) are given prescribed values required to ensure convergence (2.05, 2.05, and
.73, respectively)~\cite{clerc-2002-constricted-pso}.

@@

Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
= 0$, these extra computations will be ignored, and might just as well never
have been computed.  We call the set $\{\ofunc(\specpos_{t+1}^{c}) : \casegen
\in \caseset\}$ ``speculative children'' because only one of them will be
needed.

@@

In order to do two iterations at once, we use 8 times as many processors as
there are particles in the swarm.  If these processors were not performing
speculative evaluation, they might instead be used for function evaluation
needed to support a large swarm.  This raises the question of whether a swarm
of 100 particles doing twice as many iterations outperforms a swarm of 800
particles.  We show in the rest of this paper that in many instances a smaller
swarm performing more iterations does in fact outperform a larger swarm.  We
acknowledge that both intuition and prior
research~\cite{mcnabb-2009-large-particle-swarms} indicate that the
optimization of deceptive functions benefits greatly from large and even very
large swarm sizes. Thus this work will focus on less deceptive functions.

@@

A particle at iteration $t-1$ that has been moved to iteration $t$ using
\eqref{eq:velupdate} and \eqref{eq:posupdate}, but whose position has not yet
been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has been
evaluated, but it has still not yet received information from its neighbors, it
is denoted as $\nonbest{\p}_t$.  Only when the particle has updated its
neighborhood best is it a complete particle at iteration $t$.  It is then
simply denoted as $\p_t$.

@@

\subsection{Using All Speculative Evaluations}
\label{sec:extra}

In performing speculative evaluation as we have described it, $2n+1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of the information
obtained through those evaluations instead of ignoring it.  Making use of this
information changes the behavior of PSO, instead of reproducing it exactly as
the above method explains, but the change turns out to be an improvement in our
context.

To make better use of the speculative evaluations, instead of choosing the
speculative child that matches that branch that the original PSO would have
taken, we take the child that has the best value.  The methodology is exactly
the same as above except for the process of choosing which speculative child to
accept.  The only change needed in \alg{centralized} is in step 7, where the
$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
instead of with the matching branch. We call this variant ``Pick Best''.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations.''  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of at least seconds (or more) to
evaluate; in such cases our framework is reasonable.

In each set of experiments we keep the number of processors for each algorithm
constant.  In all of our experiments SEPSO uses topologies in which each
particle has two neighbors in addition to itself.  As shown in
Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have a total of $8p$ evaluations for every two iterations, where
$p$ is the number of particles in the speculative swarm.  In order to use the
same resources for each algorithm, we compare swarms of size $p$ in the
speculative algorithms with swarms of size $8p$ in standard PSO.

As discussed in \sec{implementation}, where the Complete topology would
normally be used, we use a Random topology in our speculative algorithm, as
Complete leads to an explosion of speculative evaluations.  For the standard
PSO baseline we have included experiments with the Ring and the Random
topologies, both with two neighbors, as well as for the Complete topology.  It
is important to note however, that in many cases the Complete topology is not a
practical alternative in the context of large swarms on large clusters where
where the messaging complexity is $O(p^2)$ and can overwhelm the system.

We follow the experimental setup used
in~\cite{bratton-2007-defining-a-standard-for-pso}.  All functions have 20
dimensions, and all results shown are averages over 20 runs.  For ease of
labeling, we call our speculative algorithm SEPSO, the variant SEPSO Pick Best,
and standard PSO just PSO and identify the topology in a suffix, ``PSO Ring''
for example.

The benchmark function Sphere ($f(\Vec{x}) = \sum_{i=1}^D x_i^2$) has no local
optima and is most efficiently optimized using a small swarm but many
iterations.  In our experiments with Sphere, we use 240 processors; thus 30
particles for SEPSO and 240 for PSO.  In \fig{sphere}, we can see that SEPSO
clearly beats PSO with a Random topology or a Ring topology.  SEPSO approaches
the performance of PSO with the Complete topology, even though PSO with the
Complete topology requires $O(p^2)$ communication.  SEPSO Pick Best handily
outperforms all other algorithms in this problem.

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  We show 10th and 90th percentiles every 100 iterations.  Note
  that PSO Complete requires $O(p^2)$ messaging and may not be practical in
  many cases.}
  \label{fig:sphere}
\end{figure}

The benchmark function Griewank is defined by the equation $f(\Vec{x}) =
\frac{1}{4000}$ $\sum_{i=1}^D x_i^2 - \Pi_{i=1}^D
\cos\left(\frac{x_i}{\sqrt{i}} \right) + 1$.  It is best solved in PSO using
the Ring topology, as Complete is prone to premature convergence on local
optima.  Griewank has a global optimum with a value of 0, and sometimes the
swarm finds the optimum and sometimes it does not.  Instead of showing average
function value at each iteration, a more enlightening plot for Griewank shows
the percent of runs that have found the global optimum by each iteration.

PSO and SEPSO get caught in local minima with small swarm sizes so we show
results in \fig{griewank} for swarms of size 100 (SEPSO) and 800 (standard)
using the Ring topology. \fig{griewank} shows that SEPSO quickly finds the
global optimum, between two and three times faster than running standard PSO.

In \fig{griewank} we also show results for the Bohachevsky function, defined as
$f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
x_{i+1})+.7)$.  Bohachevsky is a unimodal function best optimized with a
Complete swarm.  It is similar to Griewank in that there is a global optimum
with a value of 0, and swarms either find the optimum or get stuck.  Both SEPSO
algorithms find the optimum much faster than PSO Random, though only SEPSO Pick
Best beats PSO Complete.  Also, while the smaller swarm size of SEPSO gets
stuck 75\% of the time, when using SEPSO Pick Best with the same swarm size,
the algorithm finds the optimum every time.

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{griewank}
  \includegraphics[width=.45\columnwidth]{bohachevsky}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations, and function Bohachevsky with a swarm that uses 480 processors
  per round of evaluations.}
  \label{fig:griewank}
\end{figure}

Previous work has shown that the optimization of highly deceptive functions
like Rastrigin ($f(\Vec{x}) = \sum_{i=1}^D \left(x_i^2 - 10\cos\left(2\pi
x_i\right) + 10\right)$) benefit greatly from the addition of particles.
Smaller swarms get caught in local optima, up to swarms of at least 4000
particles~\cite{mcnabb-2009-large-particle-swarms}.  Because our speculative
algorithms have a significantly smaller swarm size, they get stuck at higher
values while the larger swarms performing regular PSO continue to improve the
best value found.  Our experiments with SEPSO on Rastrigin were predicably lack
luster, yielding an average value of 31 after 1000 evaluations, as compared to
10 for standard PSO.

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  Using our modifications, the original PSO algorithm is exactly
reproduced two iterations at a time.  This technique requires more function
evaluations per iteration than regular PSO, but for some functions still
performs better when run in a parallel environment.  We have also described a
method for making use of extra speculative evaluations that performs very well
on some functions.

There are some functions for which very little exploration needs to be done;
Sphere is an example of such a function.  For such functions the best use of
processors is to have a small swarm performing speculative evaluation with our
Pick Best method, where all speculative evaluations are used.

There are other functions for which it seems there is never enough exploration,
such as the Rastrigin function.  It has been shown that up to 4000 particles
there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-2009-large-particle-swarms}.  With such functions, the
smaller swarm size required by speculative evaluation is not able to produce
enough exploration to perform better than standard PSO.

Griewank and Bohachevsky are functions between Sphere and Rastrigin.  They are
deceptive and prone to premature convergence, but by adding particles to the
swarm a point is reached where ``enough'' exploration is done, and the algorithm finds the optimum essentially all of the time.  For such functions, the best approach seems to be to increase the swarm size until ``enough'' exploration is reached, then use extra processors to perform speculative evaluation and increase the number of iterations performed.  Sphere and Rastrigin can be thought of as special cases fo these types of functions; Sphere simply needs a very small swarm size to produce ``enough'' exploration, and Rastrigin requires a very large swarm.  We expect that for all functions there is a swarm size for which additional particles are less useful than additional iterations.

Large parallel clusters are often required to successfully optimize practical modern problems.  To properly use PSO with such clusters, a balance needs to be made between using processors to increase the swarm size and using them to increase the speed of the algorithm.  This work is a first step in that direction.
